---
title: "Lab 2 - Data wrangling"
author: "Prabhjot Singh"
date: "12 Feb, 2026"
format: 
  html: default # for speed
  pdf: default # for final output

---


```{r}
#| eval: true
#| message: false

library(tidyverse)
```

```{r}
midwest
```

# Questions

## Part 1

### Question 1
```{r}
#| eval: true
#| message: false

midwest|>count(state)|>arrange(desc(n))
```
#### Illinois has the highest number of counties (102), and Wisconsin has lowest number of counties (72)

\newpage

### Question 2
```{r}
midwest |>
  count(county, state) |>
  count(county, name = "freq") |>
  filter(freq >= 2)
```
\newpage

### Question 3
```{r}
midwest |>filter(popdensity > 25000) |>select(county, state, popdensity, poptotal, area)|>arrange(desc(popdensity))

```
```{r}
midwest |>
  filter(popdensity == max(popdensity)) |>
  select(county, state, popdensity, poptotal, area)


```
\newpage
### Question 4
```{r}
midwest|>summarize(median(popdensity),IQR1=quantile(popdensity,0.25),IQR3=quantile(popdensity,0.75))
```

The distribution of population density of counties is unimodal and extremely right-skewed. A typical Midwestern county has population density of 1156 people per unit area. The middle 50% of the counties have population densities between 622 to 2330 people per unit area.
\newpage

### Question 5
```{r}
#| echo: false
#| message: true
midwest_ <- midwest |>mutate(metro = if_else(inmetro == 1, "Yes", "No"))|> group_by(state) |> summarise(
    metro_popln = sum(metro == "Yes"),
    not_metro_popln = sum(metro == "No"),
    total_popln = n(),
    metro_proportion = metro_popln / total_popln,
    not_metro_proportion = not_metro_popln / total_popln
) |> select(state, metro_proportion, not_metro_proportion)

midwest_
```

\newpage

### Question 6
A
```{r}
midwest |>
    filter (percollege < 10 & percbelowpoverty > 48) |>
    select (county, state, percbelowpoverty, percollege)
```

B
```{r}
midwest |>
    filter (percollege >40 & percbelowpoverty < 17) |>
    select (county, state, percbelowpoverty, percollege)
```
C
```{r}
midwest |>
    filter ((percollege < 10 &
                 percbelowpoverty > 48) | (percollege > 40 &
                                               percbelowpoverty < 17)) |>
    select (county, state, percbelowpoverty, percollege)
```
D
```{r}
midwest |>
    mutate(potential_outlier = if_else(
        (percollege < 10 &
             percbelowpoverty > 48) | (percollege > 40 &
                                           percbelowpoverty < 17),
        "Yes",
        "No"
    )) |>
    select(county, state, percbelowpoverty, percollege, potential_outlier) |>
    arrange(potential_outlier)
```
E
```{r}
midwest |>
    mutate(potential_outlier = if_else(
        (percollege < 10 &
             percbelowpoverty > 48) | (percollege > 40 &
                                           percbelowpoverty < 17),
        "Yes",
        "No"
    )) |> ggplot(aes(
        x = percollege,
        y = percbelowpoverty,
        colour = potential_outlier,
        shape = state
    )) + geom_point() + labs(title = "Population below the Poverty Line by Population with College Degree",subtitle = "by percentage", x = "Population with College Degree (%)", y = "Population below the Poverty Line (%)", )
```

\newpage

### Question 7
A
```{r}
state_population <- midwest |>
    group_by(state) |>
    summarise(total_popln = sum(poptotal)) |>
    arrange(desc(total_popln))
state_population
```
B
```{r}
state_population|>mutate(popln_proportion=total_popln/sum(total_popln))|>arrange(desc(popln_proportion))
```

C
Illinois has the biggest proportion (27%) of people living there, Wisconsin has the smallest proportion(11%) of people living there

\newpage

### Question 8
```{r}
state_poverty<-midwest|>group_by(state)|>summarise(mean_percbelowpoverty=mean(percbelowpoverty))
state_poverty|>arrange(mean_percbelowpoverty)
```
Indiana has the lowest average percentage(10.31%) living below poverty and Michigan has the highest average percentage(14.22%) living below poverty.

\newpage

## Part 2


### Question 9

